An overview on temporal databases.\\ New approaches and proposals.

### 2012-01-17 book/sources/timeDomain.tex
A new version with extended definition of granularity and calendar.

\begin{svgraybox}
The basic elements in a temporal domain are the following:\\
%\begin{itemize}
%\item
\textbf{Instant}:  Is a time point over an underlying temporal domain.\\
%\item
\textbf{Time interval}: Is the time between two instants.\\
%\item
\textbf{Event}: Is an instantaneous fact. Something that happens at an instant.\\
%\item
\textbf{Chronon}: Is a non-decomposable time interval of fixed duration. It is possible for a temporal model to leave the chronon duration unspecified. The duration will be specified later in the implementation of the model.
%\end{itemize}
\end{svgraybox}
Any temporal model may represent the time as points as well as intervals. The equivalence between a interval-based and point-based temporal data models is demonstrated in \cite{Böhlen_point-versus}.

Attending to the density between the chronons, a model can be classified into the following three types:

\begin{itemize}
\item
\textbf{Discrete model}:  The discrete model of time is isomorphic in relation to the natural numbers. According to this model, each natural number corresponds to a \emph{chronon}. 
\item
\textbf{Dense model}: This model is isomorphic in relation to the rational or the real numbers. It represents one instant of time in the gap between another two instants.
\item
\textbf{Continuous model}: This is an extension of the dense model with no gaps. It is isomorphic in relation to the real numbers.
\end{itemize}

@@

%%% granularity
%\subsubsection{Granularity}
Next to the inherent difficulties of time, an issue is \textbf{temporal granularity}. A temporal granularity is a partitioning of the time line used by a system, usually dependent on the application. For example, the age of an adult human being is usually expressed in years: one will use sentences like \emph{`Laura is 21 years old'} instead of \emph{`Laura is 21 years, 3 months and 4 days old'}. In this example any time period shorter than a year needs no representation and thus the used granularity allows no specification for time periods shorter than a year.

@@

\subsection{An example of temporal domain: Julian Day Number}
% proposal for the underlying Julian Day Number domain.

@@

%\subsubsection{Conversion among dates}
%\paragraph{Conversion between Gregorian and Julian dates}
%The Julian Date consist on the Julian Day number and the time. It is computed with the formula \cite{Usn}:\\
%\begin{equation}
%\label{eq:julian day}
%JD = JDN + \frac{hour-12}{24} + \frac{minutes}{1440} + \frac{seconds}{86400}
%\end{equation}
%
%Here the JDN is the Julian Day Number. This equation does not take into account the leap seconds \footnote{A leap second(s) is the amount of second(s) added or deleted every year in the Universal Time.\cite{Wik},\cite{Lea}. }.

%\paragraph{Gregorian calendar to Julian day}
%Any date from Gregorian calendar \footnote{As Gregorian calendar entered in use in 1582, any use before that date is called \emph{proleptic calendar}.} can be converted into a Julian Day using the following formula:\\
%\begin{equation}
%\label{eq:JDN}
%JDN = \frac{1461x(Y + 4800 + \frac{M-14}{12})}{4}+\frac{367x(M-2-12x\frac{M-14}{12})}{12}-\frac{3x\frac{Y+4900+\frac{M-14}{12}}{100}}{4} 
%\end{equation}
%
%This formula was created by Henry Fliegel and Thomas C. van Flandern in 1968. \cite{Fliegel:1968:LEM:364096.364097} It takes into account leap years and that they are applicable only when the century is divisible by 400. The value $Y$ represents the year and the value $M$ represents the month.
%\paragraph{Julian calendar to Julian day}
%The algorithm to convert between Julian calendar and Julian Day Number is only valid for dates since 4713 B.C (positive values of Julian Day) \cite{Dog}:\\
%\begin{equation}
%\label{eq:JDN_2_JD}
%JDN=367Y-\frac{7 (Y + 5001 + \frac{M-9}{7})}{4}+\frac{275xM}{9}+D+1729777
%\end{equation}
%
%Where $Y$ is the year value, $M$ is the month value, and $D$ is the day value.
%
%%\subsubsection{Membership degree computation}
%The main advantage from the Julian day number is the ability to make calculations. Thus, the membership degree computation for a fuzzy number based on a Julian day is explained here. \\
%Let a trapezoidal possibility distribution defined by four JDN named $a$,$b$,$c$ and $d$. The fuzzy validity period \emph{FVP}, can be represented as in figure \ref{fig:fuzzy-validity-period}. The following equation shows the membership degree for a Julian day number \emph{JDN} in a \emph{FVP} represented by means of the possibility distribution of $[a,b,c,d]$. In a valid-time database, the values for the valid starting time \emph{VST} and the valid ending time \emph{VET} can be fuzzyfied as shown in equation \ref{eq:fuzzy-validity-period}.
%
%The membership function for a given $x$ number in \emph{JDN} format is the following:\\
%\begin{equation}
%\label{eq:trapezoidal distribution}
%%\[
%T(x)=
%\begin{cases}
%0, \mbox(if x \leq a, x \geq d)\\
%1, \mbox(if x \in [b,c])\\
%\frac{x-a}{b-a}   \mbox(if x \in (a,b])\\
%\frac{d-x}{d-c} 	 \mbox(if x \in (c,d])\\
%\end{cases}
%%\]
%\end{equation}
%The points outside the trapezium or inside the interval $[a,b]$ are trivial. The points inside the other intervals has the following particularities:\\
%The subtraction operation with Julian day numbers gives as result a scalar value (as explained before). The division operation between two scalars gives as result an scalar in the interval $[0,1]$ which represents the membership degree.

>>>>

\begin{svgraybox}
The basic elements in a temporal domain are the following:\\
%\begin{itemize}
%\item
\textbf{Instant}:  Is a time point over an underlying temporal domain.\\
%\item
\textbf{Time interval}: Is the time between two instants.\\
%\item
\textbf{Event}: Is an instantaneous fact. Something that happens at an instant.\\
%\item
\textbf{Chronon}: Is a non-decomposable time interval of fixed duration. It is possible for a temporal model to leave the chronon duration unspecified. The duration will be specified later in the implementation of the model.
%\end{itemize}
\end{svgraybox}

\begin{definition}
\label{def:set-chronon}
The \textbf{\emph{set of all the chronons}} in a system is denoted by $\C$. A chronon $c$ is therefore defined as:
\begin{equation}
\left \lbrace c \mid c \in \C \right \rbrace
\end{equation}
\end{definition}

Any temporal model may represent the time as points as well as intervals. The equivalence between a interval-based and point-based temporal data models is demonstrated in \cite{Böhlen_point-versus}.

Attending to the density in the set $\C$ of the chronons, a model can be classified into the following three types:

\begin{itemize}
\item
\textbf{Discrete model} ($\C \subset \Nat$):  The discrete model of time is isomorphic in relation to the natural numbers. According to this model, each natural number corresponds to a \emph{chronon}.
\item
\textbf{Dense model} ($\C \subset \Q$ or $\C \subset \R$): This model is isomorphic in relation to the rational or the real numbers. It represents one instant of time in the gap between another two instants. 
\item
\textbf{Continuous model} ($\C \subset \R$): This is an extension of the dense model with no gaps. It is isomorphic in relation to the real numbers. 
\end{itemize}

@@

%%% granularity
\subsection{\label{subsec:granularity}Granularity}
Next to the inherent difficulties of time, an issue is \textbf{temporal granularity}. A temporal granularity is a partitioning of the time line used by a system, usually dependent on the application. For example, the age of an adult human being is usually expressed in years: one will use sentences like \emph{`Laura is 21 years old'} instead of \emph{`Laura is 21 years, 3 months and 4 days old'}. In this example any time period shorter than a year needs no representation and thus the used granularity allows no specification for time periods shorter than a year.
The formal definition for granularity is given in ~\cite{Lin97}:

\begin{definition}
\label{def:granularity}
A \textbf{\emph{granularity}} $\alpha$ is an ordered set of non-overlapping and continuous time elements called granules. \\
\end{definition}

\begin{definition}
\label{def:granule}
A \textbf{\emph{granule}} is the basic time unit represented by a granularity. E.g. in the hours granularity, an hour is a granule.
As granularity $\alpha$ it is an ordered set, each \emph{granule} may be indexed by an integer number:\\
\begin{equation}
\alpha = \left \lbrace \ldots, 0_\alpha, 1_\alpha, \ldots \right \rbrace
\end{equation}
\end{definition}

In a system, the smallest granularity is the \emph{chronons} granularity, denoted with the index $\bot$. It is possible to define a function $f$ that maps from  a given granule $i_\alpha$ (in the given granularity $\alpha$) to the set of the corresponding chronons.

\begin{definition}
\label{def:mapping-function}
The \textbf{\emph{mapping function}} $f$ in a given granule $i_\alpha$ is denoted by:\\
%\begin{align}
$f:i_\alpha \rightarrow \C$\\
$f(i_\alpha) = \lbrace c_\bot \mid c_\bot$ is in $i_\alpha \rbrace \wedge c_\bot \in \C$\\
%\end{align}
\end{definition}

If $\alpha$ is a granularity, then the following properties must hold:\\
\begin{itemize}
\item
$\alpha$ is an ordered set: if $c_\bot \in f(i_\alpha)$ and $c^{'}_\bot \in f(j_\alpha)$ and $c_\bot < c^{'}_\bot$ implies that $i_\alpha < j_\alpha$.
\begin{figure}
\centering
\input{graphs/granularity-prop1.tex}
\caption{$\alpha$ is an ordered set}
\label{fig:granularity-prop1}
\end{figure}
\item
$\alpha$ is a set of continuous granules. Consider the chronons $c_\bot < c^{'}_\bot < c^{''}_\bot$. Then $c_\bot \in f(i_\alpha)$,$c^{''}_\bot \in f(i_\alpha)$ implies that $c^{'}_\bot \in f(i_\alpha)$.
\begin{figure}
\centering
\input{graphs/granularity-prop2.tex}
\caption{$\alpha$ is a set of continuous granules}
\label{fig:granularity-prop2}
\end{figure}
\item
The granules in $\alpha$ do not overlap: If $i_\alpha \neq j_\alpha$ then $f(i_\alpha) \cap f(j_\alpha) = \emptyset$.
\begin{figure}
\centering
\input{graphs/granularity-prop3.tex}
\caption{The granules in $\alpha$ do not overlap}
\label{fig:granularity-prop3}
\end{figure}
\end{itemize}

The mapping function $f$ allows the comparison among granularities:

\begin{definition}
\label{def:finner-than}
A granularity $\alpha$ is \emph{``finner than''} granularity $\beta$ when the set of corresponding chronons for a given granule $i_\alpha$ is smaller than the set of the corresponding chronons for a given granule $j_\beta$.
\begin{equation}
\label{eq:finner-than}
\mid f \left( i_\alpha \right) \mid < \mid  f \left( j_\beta \right) \mid
\end{equation} 
\end{definition}

\begin{definition}
\label{def:coarser-than}
A granularity $\alpha$ is \emph{``coarser than''} granularity $\beta$ when the set of corresponding chronons for a given granule $i_\alpha$ is greater than the set of the corresponding chronons for a given granule $j_\beta$.
\begin{equation}
\label{eq:coarser-than}
\mid f \left( i_\alpha \right) \mid > \mid  f \left( j_\beta \right) \mid
\end{equation} 
\end{definition}

A function,\emph{Cast} may be specified to map between two granularities $\alpha$ and $\beta$. The specification for that function is:
\begin{equation}
\label{eq:cast-function}
\mbox{Cast} \left( i_\alpha,\alpha,\beta \right) = j_\beta
\end{equation}
This function associates a granule $i_\alpha$ in the granularity $\alpha$ to a corresponding granule $j_\beta$ in the corresponding granularity $\beta$. An \emph{upwards mapping} is the mapping from a granularity $\alpha$ to a coarser granularity $\beta$, whereas a downwards mapping is the mapping from a granularity $\gamma$ to a finner granularity $\delta$. Figure \ref{fig:granularity-graph-example} shows a granularity graph example with the Gregorian Calendar and the academic days as possible granularities.

\begin{figure}
\centering
\includegraphics[scale=0.5]{graphs/granularityGraph.pdf}
\caption{A granularity graph example}
\label{fig:granularity-graph-example}
\end{figure}

The mapping between two granularities may be classified into the following types and are explained within the figure \ref{fig:granularity-graph-example}:
\begin{itemize}
\item
Regular mapping: When the mapping function between two granularities is calculated by means of multiplications or divisions and maybe some anchor adjustment. E.g. the mapping between hours and minutes. This mapping is represented in the figure with a normal arrow.
\item
Irregular mapping: When the mapping function can not be calculated by means of multiplications or divisions. E.g. the mapping between months and days. (This mapping is represented in the figure with a bold arrow.
\item
Congruent mapping: Two granularities that have the same granules but with different anchos. E.g. the mapping between the Gregorian Calendar and academic days. This mapping is represented in the figure with a straight line without arrow.
\end{itemize}

@@

\subsection{\label{subsec:timedomain-calendar}Time domain: Calendars}
%maybe introduce previously the concept of time units, linear hierarchy and time point.
A Calendar is a organization of different granularities (e.g. day, month, year) in a hierarchy. 
In order to define a calendar, some previous concepts should be introduced~\cite{Kraus1997}:

\begin{definition}
\label{def:granularity-hierarchy}
\textbf{(Linear hierarchy of granularities)} A linear hierarchy of granularities is a finite collection of granularities with a linear order $\sqsubset$ among these granularities. The coarser granularity is called the top of the hierarchy and it must be an infinite set while all the others granularities must be finite sets.
\end{definition}

The hierarchy of days $\sqsubset$ months $\sqsubset$ years defines a linear hierarchy.

\begin{definition}
\label{def:linear-calendar}
\textbf{(Linear Calendar)} A linear calendar consist of a linear hierarchy $H$ of granularities and a validity predicate which specifies if a granule is valid in  the calendar.
\end{definition}

The linear hierarchy days $\sqsubset$ months $\sqsubset$ years represents a linear calendar. It is necessary to define the validity predicates because 17 January 2012 is a valid element of the calendar whereas 30 February 2012 is not.

\subsubsection{\label{subsubsubsec:julian-day-number}An example of temporal domain: Julian Day Number}
% proposal for the underlying Julian Day Number domain.

@@
